---
title: "ENV 797 - Time Series Analysis for Energy and Environment Applications | Spring 2025"
subtitle: "Assignment 7 - Due date 03/06/25"
author: "Aye Nyein Thu"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A07_Sp25.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

## Set up

```{r package, warning=FALSE, message=FALSE}
#Load/install required package here
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(cowplot)
library(dplyr)
library(trend)

```

## Importing and processing the data set

Consider the data from the file "Net_generation_United_States_all_sectors_monthly.csv". The data corresponds to the monthly net generation from January 2001 to December 2020 by source and is provided by the US Energy Information and Administration. **You will work with the natural gas column only**.


### Q1

Import the csv file and create a time series object for natural gas. Make you sure you specify the **start=** and **frequency=** arguments. Plot the time series over time, ACF and PACF.

```{r data import and plot, warning=FALSE, message=FALSE}
# Import data
netgen <- read.csv(
  file = "./Data/Net_generation_United_States_all_sectors_monthly.csv",
  header = TRUE, skip=4)

# Prepare data
netgen_cleaned <- netgen %>%
  mutate(Month = my(Month)) %>% 
  select(Month, Natural_gas = "natural.gas.thousand.megawatthours") %>% 
  arrange(Month)

# Create time series
ts_netgen <- ts(netgen_cleaned$Natural_gas,
                           start=c(year(netgen_cleaned$Month[1]),
                                   month(netgen_cleaned$Month[1])),
                           frequency=12) 
# Plot ACF and PACF 
par(mfrow=c(1,2))

ACF_Plot <- Acf(ts_netgen, lag = 40, plot = TRUE)
PACF_Plot <- Pacf(ts_netgen, lag = 40)

# Plot ACF and PACF 
plot_grid(
  autoplot(ts_netgen), 
  autoplot(Acf(ts_netgen,lag.max=40, plot = FALSE)),  
  autoplot(Pacf(ts_netgen,lag.max=40, plot = FALSE)),
  nrow=1
)
```

### Q2

Using the $decompose()$ and the $seasadj()$ functions create a series without the seasonal component, i.e., a deseasonalized natural gas series. Plot the deseasonalized series over time and corresponding ACF and PACF. Compare with the plots obtained in Q1.

```{r deseasoned, , warning=FALSE, message=FALSE}
# Decompose the series
decompose_netgen <- decompose(ts_netgen, "additive")
  
# Deseasonalized the series 
deseason_netgen <- seasadj(decompose_netgen)
  
# Plot the series 
plot_grid(
  autoplot(deseason_netgen),
  autoplot(Acf(deseason_netgen,lag.max=40, plot = FALSE)),  
  autoplot(Pacf(deseason_netgen,lag.max=40, plot = FALSE)),
  nrow=1
)
```

> Answer: In the original series, both ACF and PACF plots exhibits the existence of seasonality. The ACF plots has the regular seasonality-like features and spikes in 12, 24 and 36 while PACF also showcases the similar spikes behaviors. After deseasoning the series, the seasonality features have now been removed as there is not regular spikes in both ACF and PACF plots after deasonalization. The ACF plot shows the trend behavior. 

## Modeling the seasonally adjusted or deseasonalized series

### Q3

Run the ADF test and Mann Kendall test on the deseasonalized data from Q2. Report and explain the results.

```{r model the deseasonalized series, warning=FALSE, message=FALSE}
# Run the ADF test
summary(MannKendall(deseason_netgen))

# Run the Mann Kendall Test 
print(adf.test(deseason_netgen, alternative = "stationary")) 
```

> Answer: The Mann Kendall test result indicates that the deseasonalized series of net natural gas generation has a very strong increasing trend at tau value of 0.84. As the p-value is very small, the result is statistically significant at 1% level. 
The Augmented Dickey-Fuller (ADF) test shows that the deseasonalized series has a ADF value at -4.03 and p-value at 0.01. Since the p-value is lower than 0.05, we have enough evidence to reject the null hypothesis that suggests the deseasonalized series is a unit root or stochastic trend. Therefore, the deseasonalized series is stationary. 

### Q4

Using the plots from Q2 and test results from Q3 identify the ARIMA model parameters $p,d$ and $q$. Note that in this case because you removed the seasonal component prior to identifying the model you don't need to worry about seasonal component. Clearly state your criteria and any additional function in R you might use. DO NOT use the $auto.arima()$ function. You will be evaluated on ability to understand the ACF/PACF plots and interpret the test results.

> Answer: Using the plots in Q2, the parameters of the manual ARIMA model would be p = 1, d = 0 and q = 0. We assume that the order of non-seasonal AR is 1 since the ACF plot shows a strong autocorrelation and significant decaying behavior; and the PACF shows a clear cut-off due to spike at lag 1 and goes down thereafter. The desesonalized series therefore is a strong AR. As we do not see any cut-off in ACF plot, we assume that q = 0. Although the deasonalized series and ACF plots show an upward trend that might not be stationary, the ADF test indicates that the series is stationary. We will take the test result and assume that q = 0 as we do not need to difference the stationary series. But it is also necessary to note down the d could also be 1. 

### Q5

Use `Arima()` from package "forecast" to fit an ARIMA model to your series considering the order estimated in Q4. You should allow constants in the model, i.e., `include.mean = TRUE` or `include.drift=TRUE`. **Print the coefficients** in your report. Hint: use the `cat()` or `print()` function to print.

```{r manual ARIMA}
manual_ARIMA <- Arima(deseason_netgen,
                       order=c(1,0,0),
                       include.drift=TRUE)

print(manual_ARIMA)
```

### Q6

Now plot the residuals of the ARIMA fit from Q5 along with residuals ACF and PACF on the same window. You may use the $checkresiduals()$ function to automatically generate the three plots. Do the residual series look like a white noise series? Why?

```{r ARIMA residuals plot}
checkresiduals(manual_ARIMA)
```

> Answer: The residuals plot of the deseasonalized series seem to be random as there is no clear trends or regular spikes. The ACF plot also shows that most of the correlation lines are within the blue line and looks random. The residuals distribution also showcases a normal distribution. Therefore, we could assume that it exhibits a white noise series. 


## Modeling the original series (with seasonality)

### Q7

Repeat Q3-Q6 for the original series (the complete series that has the seasonal component). Note that when you model the seasonal series, you need to specify the seasonal part of the ARIMA model as well, i.e., $P$, $D$ and $Q$.

```{r seasonal ARIMA}
# Run the ADF test
summary(MannKendall(ts_netgen))

# Run the Mann Kendall Test 
print(adf.test(ts_netgen, alternative = "stationary")) 

# Redo the plot for quick check 
# Plot ACF and PACF 
plot_grid(
  autoplot(ts_netgen), 
  autoplot(Acf(ts_netgen,lag.max=40, plot = FALSE)),  
  autoplot(Pacf(ts_netgen,lag.max=40, plot = FALSE)),
  nrow=1
)

# Specify the model parameters 
# Non-seasonal part 
# p = 1 (We could see a clear cut-off in the PACF plot at lag 1 and the ACF plot shows a strong long memory excluding seasonality behaviors.)
# d = 0 (We assume d = 0 as the ADF test shows it is stationary.)
# q = 0 (We assume q = 0 since there is no clear cut-off seen in ACF plot before the seasonality behavior.)

# Seasonal part 
# P = 1 (We assume P = 1 due the single spike at seasonal lags in the PACF plot.)
# D = 0 (We assume D = 0 since the ADF test is stationary.)
# Q = 0 (We assume Q = 0 due to the multiple spikes at seasonal lags in ACF plot which is the prominent feature of AR model.)
```


### Q8

Compare the residual series for Q7 and Q6. Can you tell which ARIMA model is better representing the Natural Gas Series? Is that a fair comparison? Explain your response.

```{r}

```


## Checking your model with the auto.arima()

**Please** do not change your answers for Q4 and Q7 after you ran the $auto.arima()$. It is **ok** if you didn't get all orders correctly. You will not loose points for not having the same order as the $auto.arima()$.

### Q9

Use the $auto.arima()$ command on the **deseasonalized series** to let R choose the model parameter for you. What's the order of the best ARIMA model? Does it match what you specified in Q4?

```{r}

```


### Q10

Use the $auto.arima()$ command on the **original series** to let R choose the model parameters for you. Does it match what you specified in Q7?

```{r}

```

